A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-18 tex/body.tex
more

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
%\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Ec}{E}
%\newcommand{\Ec}{\mathcal{E}}
%\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Fc}{F}
%\newcommand{\Fc}{\mathcal{F}}
%\newcommand{\Gc}{\mathsf{G}}
%\newcommand{\Ac}{\mathsf{A}}
%\newcommand{\Dc}{\mathsf{D}}
%\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\AST(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\ror}{\parallel}
%\newcommand{\ror}{{\&}}
%\newcommand{\ROR}{\kern.45em \ror \kern.45em}
\newcommand\ROR{\betweenantes}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
%\newcommand{\interp}[2]{#1\llbracket #2 \rrbracket}
%%% @ANDY: I prefer a notation that puts more emphasis on #2 (which is, in essence,
%%% the object that is returned) and less on #1
%\newcommand{\interp}[2]{\llbracket #2 \rrbracket_{#1}}
%%% And one that doesn't look so semantic
%\newcommand{\interp}[2]{\lceil #2 \rceil_{#1}}
\newcommand{\interp}[2]{\langle \smash{#2} \rangle_{#1}}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{[#1]}
\newcommand{\J}{\mathcal{J}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}
\newcommand{\ValC}{\mathcal{A}^\star}
\newcommand{\Var}{\mathcal{V}}
%\newcommand{\Varec}[1]{\Var{\ec{#1}}}
\newcommand{\Varec}[1]{\vvthinspace\widetilde{#1}\vvthinspace}
\newcommand{\VAREC}[1]{\widetilde{\vphantom{\scriptstyle x^i}\smash{#1}}}
%\newcommand{\nf}[1]{{{#1}{\downarrow}}}
\newcommand{\nf}[1]{\lfloor#1\rfloor}
\newcommand{\aequiv}{\mathrel{=_\alpha}}
\newcommand{\vsim}{\aequiv}
\newcommand{\vsimv}[1]{\mathrel{=^{#1}_\alpha}}
\newcommand{\tpos}[2]{#1\!\mid_{#2}}
\newcommand{\muvar}{\mathcal{Var}}

@@

\begin{examplex}
Let $\delta$ be a codatatype with the constructors $\const{C}, \const{D}, \const{E} :\delta \rightarrow \delta$.
Let $\Ec$ be the set
$\{
w \teq \const{C}( z ),\;
z \teq \const{D}( x ),\;
u \teq \const{C}( x ),\;
v \teq \const{E}( y ),\;
x \not\teq z
\}$.
After applying the calculus to saturation on $\Ec$, the mapping $\Val$ is as follows:
%\[\begin{array}{r@{}c@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l}
%\[\begin{array}{r@{}c@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l}
\[\begin{array}{r@{}c@{}r@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l@{\qquad}r@{}c@{}l@{\;}l}
\Val \ec{x} & = & {\Varec{x}} &
\Val \ec{z} & = & \MU {\Varec{z}}. & \multicolumn{2}{@{}l@{}}{\const{D}( {\Varec{x}} )} & &
\Val \ec{u} & {}\,=\,{} & \MU {\Varec{u}}. & \const{C}( {\Varec{x}} ) \\
\Val \ec{y} & = & {\Varec{y}} &
\Val \ec{w} & {}\,=\,{} & \MU {\Varec{w}}. & \const{C}( & \MU {\Varec{z}}. & \const{D}( {\Varec{x}} ) ) &
\Val \ec{v} & = & \MU {\Varec{v}}. & \const{E}( {\Varec{y}} )
\end{array}\]
%
To construct a completion $\ValC$, we must choose values for the free
variables of $\Val$, namely ${\Varec{x}}$ and ${\Varec{y}}$.
Modulo $\alpha$-equivalence, the set \smash{$\mathcal{T}^{\smash{\J}}_{{\Varec{x}}}( \Val )$} is
\[%\begin{array}{c}
\{
\MU x.\; \const{D}( x ),\;
\MU x.\; \const{C}( \const{D}( x )),\;
\const{C}( \MU x.\; \const{D}( x )),\;
\MU x.\; \const{C}( x )
\}
%\end{array}
\]
Consider a normal interpretation $\J$ that interprets variables in $\Ec$ based on the mapping $\Val$:
$x$ is interpreted as $\Val \ec{x}$, $y$ as $\Val \ec{y}$, and so on.
Assigning a value for $\Val \ec{x}$ that is $\alpha$-equivalent to a value in $\mathcal{T}^{\smash{\J}}_{{\Varec{x}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent,
which in turn may cause $\Ec$ to be falsified by $\J$.
For example, assign $\MU {\Varec{x}}.\; \const{D}( {\Varec{x}} )$ for ${\Varec{x}}$.
After this substitution, $\Val \ec{z}$ is $\MU {\Varec{z}}.\; \const{D}( \MU {\Varec{x}}.\; \const{D}( {\Varec{x}} ) )$,
which has normal form $\MU {\Varec{z}}.\; \const{D}( {\Varec{z}} )$,
which is $\alpha$-equivalent to $\MU {\Varec{x}}.\; \const{D}( {\Varec{x}} )$.
However, this contradicts the disequality $x \not\teq z$ in~$\Ec$.
On the other hand, if the value assigned to ${\Varec{x}}$ is fresh, %not $\alpha$-equivalent to any term in $\mathcal{T}^{\smash{\M}}_{{\Varec{x}}}( \Val )$,
the values in the range of $\Val$ remain $\alpha$-disequivalent.
For example, we may assign a value such as
$\MU {\Varec{x}}.\; \const{E}( {\Varec{x}} )$ or
$\MU {\Varec{x}}.\; \const{D}( \const{C}( {\Varec{x}} ) )$
to ${\Varec{x}}$.
Legal substitutions for ${\Varec{x}}$ may cause the range of $\Val$ to contain abnormal terms.
For example, after assigning $\MU {\Varec{x}}.\; \const{D}( \const{C}( {\Varec{x}} ) )$ to $\Varec{x}$, we have that
$\Val \ec{u}$ is $\MU {\Varec{u}}.\; \const{C}( \MU {\Varec{x}}.\; \const{D}( \const{C}( {\Varec{x}} ) ) )$,
which has normal form
$\MU {\Varec{u}}.\; \const{C}( \MU {\Varec{x}}.\; \const{D}( {\Varec{u}} ) )$.
\xend
\end{examplex}

@@

%%% TODO: Fix data

>>>>

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
%\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Ec}{E}
%\newcommand{\Ec}{\mathcal{E}}
%\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Fc}{F}
%\newcommand{\Fc}{\mathcal{F}}
%\newcommand{\Gc}{\mathsf{G}}
%\newcommand{\Ac}{\mathsf{A}}
%\newcommand{\Dc}{\mathsf{D}}
%\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\AST(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\ror}{\parallel}
%\newcommand{\ror}{{\&}}
%\newcommand{\ROR}{\kern.45em \ror \kern.45em}
\newcommand\ROR{\betweenantes}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
%\newcommand{\interp}[2]{#1\llbracket #2 \rrbracket}
%%% @ANDY: I prefer a notation that puts more emphasis on #2 (which is, in essence,
%%% the object that is returned) and less on #1
%\newcommand{\interp}[2]{\llbracket #2 \rrbracket_{#1}}
%%% And one that doesn't look so semantic
%\newcommand{\interp}[2]{\lceil #2 \rceil_{#1}}
\newcommand{\interp}[2]{\langle \smash{#2} \rangle_{#1}}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{[#1]}
\newcommand{\J}{\mathcal{J}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}
\newcommand{\ValC}{\mathcal{A}^\star}
\newcommand{\Var}{\mathcal{V}}
%\newcommand{\Varec}[1]{\Var{\ec{#1}}}
\newcommand{\Varec}[1]{\vvthinspace\widetilde{#1}\vvthinspace}
\newcommand{\VAREC}[1]{\widetilde{\vphantom{\scriptstyle x^i}\smash{#1}}}
%\newcommand{\nf}[1]{{{#1}{\downarrow}}}
\newcommand{\nf}[1]{\lfloor#1\rfloor}
\newcommand{\aequiv}{\mathrel{=_\alpha}}
\newcommand{\vsim}{\aequiv}
\newcommand{\vsimv}[1]{\mathrel{=^{#1}_\alpha}}
\newcommand{\tpos}[2]{#1\!\mid_{#2}}
%\newcommand{\muvar}{\mathcal{Var}}
\newcommand{\muvar}{\mathrm{Var}} %% cf. FV

@@

\begin{examplex}
Let $\delta$ be a codatatype with the constructors $\const{C}, \const{D}, \const{E} :\delta \rightarrow \delta$.
Let $\Ec$ be the set
$\{
w \teq \const{C}( z ),\;
z \teq \const{D}( x ),\;
u \teq \const{C}( x ),\;
v \teq \const{E}( y ),\;
x \not\teq z
\}$.
After applying the calculus to saturation on $\Ec$, the mapping $\Val$ is as follows:
%\[\begin{array}{r@{}c@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l}
%\[\begin{array}{r@{}c@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l}
\[\begin{array}{r@{}c@{}r@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l@{\qquad}r@{}c@{}l@{\;}l}
\Val \ec{x} & = & {\Varec{x}} &
\Val \ec{z} & = & \MU {\Varec{z}}. & \multicolumn{2}{@{}l@{}}{\const{D}( {\Varec{x}} )} & &
\Val \ec{u} & {}\,=\,{} & \MU {\Varec{u}}. & \const{C}( {\Varec{x}} ) \\
\Val \ec{y} & = & {\Varec{y}} &
\Val \ec{w} & {}\,=\,{} & \MU {\Varec{w}}. & \const{C}( & \MU {\Varec{z}}. & \const{D}( {\Varec{x}} ) ) &
\Val \ec{v} & = & \MU {\Varec{v}}. & \const{E}( {\Varec{y}} )
\end{array}\]
%
To construct a completion $\ValC$, we must choose values for the free
variables of $\Val$, namely ${\Varec{x}}$ and ${\Varec{y}}$.
Modulo $\alpha$-equivalence, the set \smash{$\mathcal{T}^{\smash{\J}}_{{\Varec{x}}}( \Val )$} is
$%\begin{array}{c}
\{
\MU x.\; \const{D}( x ),\;
\MU x.\; \const{C}( \const{D}( x )),\allowbreak\;
\const{C}( \MU x.\; \const{D}( x )),\;
\MU x.\; \const{C}( x )
\}
%\end{array}
$.
Consider a normal interpretation $\J$ that interprets variables in $\Ec$ based on the mapping $\Val$:
$x$ is interpreted as $\Val \ec{x}$, $y$ as $\Val \ec{y}$, and so on.
Assigning a value for $\Val \ec{x}$ that is $\alpha$-equivalent to a value in $\mathcal{T}^{\smash{\J}}_{{\Varec{x}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent,
which in turn may cause $\Ec$ to be falsified by $\J$.
For example, assign $\MU {\Varec{x}}.\; \const{D}( {\Varec{x}} )$ for ${\Varec{x}}$.
After this substitution, $\Val \ec{z}$ is $\MU {\Varec{z}}.\; \const{D}( \MU {\Varec{x}}.\; \const{D}( {\Varec{x}} ) )$,
which has normal form $\MU {\Varec{z}}.\; \const{D}( {\Varec{z}} )$,
which is $\alpha$-equivalent to $\MU {\Varec{x}}.\; \const{D}( {\Varec{x}} )$.
However, this contradicts the disequality $x \not\teq z$ in~$\Ec$.
On the other hand, if the value assigned to ${\Varec{x}}$ is fresh, %not $\alpha$-equivalent to any term in $\mathcal{T}^{\smash{\M}}_{{\Varec{x}}}( \Val )$,
the values in the range of $\Val$ remain $\alpha$-disequivalent.
For example, we may assign a value such as
$\MU {\Varec{x}}.\; \const{E}( {\Varec{x}} )$ or
$\MU {\Varec{x}}.\; \const{D}( \const{C}( {\Varec{x}} ) )$
to ${\Varec{x}}$.
Legal substitutions for ${\Varec{x}}$ may cause the range of $\Val$ to contain abnormal terms.
For example, after assigning $\MU {\Varec{x}}.\; \const{D}( \const{C}( {\Varec{x}} ) )$ to $\Varec{x}$, we have that
$\Val \ec{u}$ is $\MU {\Varec{u}}.\; \const{C}( \MU {\Varec{x}}.\; \const{D}( \const{C}( {\Varec{x}} ) ) )$,
which has normal form
$\MU {\Varec{u}}.\; \const{C}( \MU {\Varec{x}}.\; \const{D}( {\Varec{u}} ) )$.
\xend
\end{examplex}

@@
